Safe Transferable Regions

### 2018-06-05 paper/abstract.tex
Edits.
Typos etc fixed.

There is an increasing interest in alternative memory management
schemes that seek to the combine the convenience of garbage collection
and the performance of manual memory management in a single language
framework.  Unfortunately, ensuring safety in presence of manual
memory management remains as great a challenge as ever. In this paper,
we present a C\#-like object-oriented language called \name that uses
a combination of region type system and lightweight runtime checks to
enforce safety in presence of user-managed memory regions called
\emph{transferable regions}. Unsafe transferable regions have been
previously used in dataflow computations to contain the latency due to
unbounded GC pauses. Our approach shows that it is possible to restore
safety without compromising on the benefits of transferable regions.
We prove the type safety of \name in a formal framework that includes
its C\#-inspired features, such as higher-order functions and
generics. We complement our type system with a type inference
algorithm, which eliminates the need for programmers to write region
annotations on types. The inference algorithm has been proven sound
and relatively complete. We describe a prototype implementation of the
inference algorithm, and our experience of using it to enforce memory
safety in dataflow programs.
% %
% In this paper, we present a language and a memory management scheme based
% on user-managed memory regions (called \emph{transferable regions})
% that allow programmers to exploit knowledge of data structures' lifetimes
% to improve performance.

>>>>

There is an increasing interest in alternative memory management
schemes that seek to combine the convenience of garbage collection
and the performance of manual memory management in a single language
framework.  Unfortunately, ensuring safety in presence of manual
memory management remains as great a challenge as ever. In this paper,
we present a C\#-like object-oriented language called \name that uses
a combination of region type system and lightweight runtime checks to
enforce safety in presence of user-managed memory regions called
\emph{transferable regions}. Unsafe transferable regions have been
previously used to contain the latency due to unbounded GC pauses. Our
approach shows that it is possible to restore safety without
compromising on the benefits of transferable regions.  We prove the
type safety of \name in a formal framework that includes its
C\#-inspired features, such as higher-order functions and generics. We
complement our type system with a type inference algorithm, which
eliminates the need for programmers to write region annotations on
types. The inference algorithm has been proven sound and relatively
complete. We describe a prototype implementation of the inference
algorithm, and our experience of using it to enforce memory safety in
dataflow programs.
% %
% In this paper, we present a language and a memory management scheme based
% on user-managed memory regions (called \emph{transferable regions})
% that allow programmers to exploit knowledge of data structures' lifetimes
% to improve performance.

### 2018-06-05 paper/overview.tex
Typos etc fixed.

Fig~\ref{fig:motivating-eg-in-broom} shows how the motivating example
of Fig.~\ref{fig:motivating-eg} can be written in \name.
The \C{onReceive} method receives its input message in a \emph{transferred}
region (\ie, a \emph{closed} region whose ownership is transferred to the
recipient).
Line 7 creates a new region to store the output for time \C{t},
initializing it to contain an empty list.
Line 9 opens the input region to process it.
Line 10 creates a stack region \C{R0}.
Thus, the temporary objects created by the iteration in line 11,
for example, will be allocated in this stack region that lives just
long enough.
We open the desired output region in line 12, so that the new output
objects created by the invocation of \C{selector} in line 13
are allocated in the output region.
Finally, the input region is freed in line 19.

>>>>

Fig~\ref{fig:motivating-eg-in-broom} shows how the motivating example
of Fig.~\ref{fig:motivating-eg} can be written in \name. The
\C{onReceive} method receives its input message in a
\emph{transferred} region (\ie, a \emph{closed} region whose ownership
is transferred to the recipient). On Line 7, we create a new region to
store the output for time \C{t}, initializing it to contain an empty
list. On Lines 8 and 9, we open the input region \C{inRgn} followed by
creating a new stack region \C{R0}. The temporary objects created by
the iteration on line 10, for example, will be allocated in the stack
region \C{Ro} that lives just long enough.  We open the desired output
region on line 11, so that the new output objects created by the
invocation of \C{selector} on line 12 are allocated in the output
region.  Finally, the input region is freed on line 14.

### 2018-06-05 paper/overview2.tex
Typos etc fixed.

\mypara{Motivating Example}
% We now illustrate how the features of \name can be used to 
Fig.~\ref{fig:motivating-eg-in-broom} shows how the motivating example
of Fig.~\ref{fig:motivating-eg} can be written in \name.  The
\C{onReceive} method receives its input message in a
\emph{transferred} region (\ie, a \emph{closed} region whose ownership
is transferred to the recipient).  Line 7 creates a new region to
store the output for time \C{t}, initializing it to contain an empty
list.  Line 8 opens the input region to process it.
% \footnote{We omit \C{@R} annotation in \C{open} when we don't need \C{R}.}
Line 9 creates a stack region \C{R0}.  Thus, the temporary objects created by
the iteration in line 10, for example, will be allocated in this stack
region that lives just long enough.  We open the desired output region
in line 11, so that the new output objects created by the invocation
of \C{selector} in line 12 are allocated in the output region.
Finally, the input region is freed in line 14. The output region at
\C{map[t]} stays as along as input messages with timestamp \C{t} keep
arriving. When the timing message for \C{t} arrives, the \C{onNotify}
method transfers the \C{outRgn} at \C{map[t]} to a downstream actor.
% After the transfer, \C{SelectVertex} is no longer allowed to access
% the \C{outRgn}, facilitating its deallocation.

@@

\mypara{Cloning} Note that in the example from
Fig.~\ref{fig:motivating-eg-in-broom} the object returned by the
\C{selector} (on Line 13) should not contain any references to the
input object, since the input region, where the object resides, will
be freed at the end of the method. If there is a need for the output
object to point to subobjects of the input object, such subobjects
must be cloned (to copy them from the input region to the output
region).  Fortunately, \name's region type system
(\S~\ref{sec:type-system}) is capable of capturing such nuances in the
type of \C{selector} and the type checker will ensure correctness.
Furthermore, the type can be automatically inferred by \name's region
type inference (\S~\ref{sec:type-inference}), which can perform the
above reasoning on behalf of the programmer.

>>>>

\mypara{Motivating Example}
% We now illustrate how the features of \name can be used to 
Fig~\ref{fig:motivating-eg-in-broom} shows how the motivating example
of Fig.~\ref{fig:motivating-eg} can be written in \name. The
\C{onReceive} method receives its input message in a
\emph{transferred} region (\ie, a \emph{closed} region whose ownership
is transferred to the recipient). On Line 7, we create a new region to
store the output for time \C{t}, initializing it to contain an empty
list. On Lines 8 and 9, we open the input region \C{inRgn} followed by
creating a new stack region \C{R0}. The temporary objects created by
the iteration on line 10, for example, will be allocated in the stack
region \C{R0} that lives just long enough.  We open the desired output
region on line 11, so that the new output objects created by the
invocation of \C{selector} on line 12 are allocated in the output
region.  Finally, the input region is freed on line 14. The output
region at \C{map[t]} stays as along as input messages with timestamp
\C{t} keep arriving. When the timing message for \C{t} arrives, the
\C{onNotify} method transfers the \C{outRgn} at \C{map[t]} to a
downstream actor.
% After the transfer, \C{SelectVertex} is no longer allowed to access
% the \C{outRgn}, facilitating its deallocation.

@@

\mypara{Cloning} Note that in the example from
Fig.~\ref{fig:motivating-eg-in-broom} the object returned by the
\C{selector} (on Line 12) should not contain any references to the
input object, since the input region, where the object resides, will
be freed at the end of the method. If there is a need for the output
object to point to subobjects of the input object, such subobjects
must be cloned (to copy them from the input region to the output
region).  Fortunately, \name's region type system
(\S~\ref{sec:type-system}) is capable of capturing such nuances in the
type of \C{selector} and the type checker will ensure correctness.
Furthermore, the type can be automatically inferred by \name's region
type inference (\S~\ref{sec:type-inference}), which can perform the
above reasoning on behalf of the programmer.

### 2018-06-05 paper/type-system.tex
Typos etc fixed.

letregion $\pi_0$ in
  let snd = new Object<$\pi_0$>() in
  letregion $\pi_1$ in
    let fst = new Object<$\pi_1$>() in
    let p = new Pair<Object,Object><$\pi_1$,$\pi_1$,$\pi_0$> (fst,snd);
\end{codejava}
%\vspace*{-0.1in}
In the above code, the instantiation of $\rho_0$ and $\rho_1$ with
$\pi_1$, and $\rho_2$ with $\pi_0$ is allowed because (a) $\pi_0$ and
$\pi_1$ are live during the instantiation, and (b). $\pi_0 \outlives
\pi_1$ and $\pi_1 \outlives \pi_1$ (since outlives is reflexive).
Observe that the region type of \C{p} conveys the fact that (a). it is
allocated in region $\pi_1$, and (b). it holds references to objects
allocated in region $\pi_0$ and $\pi_1$.  In contrast, if we choose to
allocate the \C{snd} object also in $\pi_1$, then \C{p} would be
contained in $\pi_1$, and its region type would be
\C{Pair<\ObjZ,\ObjZ><$\pi_1$,$\pi_1$,$\pi_1$>}, which we abbreviate as
\C{Pair<\ObjZ,\ObjZ>@$\pi_1$}. In general, we treat
$B\inang{\tbar}@\pi$ as being equivalent to
$B\inang{\tbar}\inang{\rbar}$. Region annotation on type $a$, where
$a$ is a type variable, assumes the form $a@\pi$. If $a$ is
instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
\C{Pair} object contained in  $\pi$.

Classes in \FB are independently parameterized over types and regions.
While this design decision has a downside that it allows type
variables to only denote the types of objects contained in a single
region, it yields benefits that outweigh the costs. In particular, it
lets us support region-polymorphic higher-order functions as class
fields.  This allows, for example, a generic class, whose type
parameters are $a$ and $b$, to contain a region-polymorphic function
of type $\inang{\rho_0,\rho_1,\rho_2 \,|\, \rho_1 \outlives \rho_0
\conj \rho_2 \outlives \rho_0}(a@\rho_1,b@\rho_2) \rightarrow
\C{Pair}\inang{a,b}\inang{\rho_0,\rho_1,\rho_2}$ as its field. Such
region-polymorphic higher-order fields are used frequently by the
dataflow operators, which apply them in the context of various regions
(\emph{e.g.,} see Fig.~\ref{fig:motivating-eg-in-broom}). Keeping type
and region parameterizations separate also simplifies the type system
so that inference becomes practical (Sec.~\ref{sec:type-inference}).
The need for polymorphism at the field level is also why \FB treats
function closures specially, rather than as objects of type \C{Func}
as in C\#.

@@

@@

The operational semantics of \FB describes a non-trivial runtime
component that introduces memory regions and performs run-time
verification of their typestate. Fig.~\ref{fig:fb-syntax-extra} shows the
additional language constructs of \FB that manifest only at run-time.
A location ($\loc$) abstracts all the memory locations associated with
a region (i.e., each memory region is associated with a single
location). A transferable region can be in one of three possible
states: closed ($\USED$), open/live ($\LIVE$), and transferred/freed
($\XFERRED$). Constructs \C{letd} and \C{opened} are the run-time
manifestations of \C{let} and \C{open}; \C{let} reduces to \C{letd}
while allocating a (static) region, and \C{open} reduces to \C{opened}
while opening a (transferable) region. An \C{opened}
expression is tagged with the typestate ($s$) of the transferable
region before it is opened (as Fig.~\ref{fig:region-fsm} shows, a
transferable region can be \C{open}'d from either an open state or a
closed state). Special value $\bot$ denotes an exception.

>>>>

letregion $\pi_0$ in
  let snd = new Object<$\pi_0$>() in
  letregion $\pi_1$ in
    let fst = new Object<$\pi_1$>() in
    let p = new Pair<Object,Object><$\pi_1$,$\pi_1$,$\pi_0$> (fst,snd);
\end{codejava}
%\vspace*{-0.1in}
In the above code, the instantiation of $\rho_0$ and $\rho_1$ with
$\pi_1$, and $\rho_2$ with $\pi_0$ is allowed because (a) $\pi_0$ and
$\pi_1$ are live during the instantiation, and (b). $\pi_0 \outlives
\pi_1$ and $\pi_1 \outlives \pi_1$ (since outlives is reflexive).
Observe that the region type of \C{p} conveys the fact that (a). it is
allocated in region $\pi_1$, and (b). it holds references to objects
allocated in region $\pi_0$ and $\pi_1$.  In contrast, if we choose to
allocate the \C{snd} object also in $\pi_1$, then \C{p} would be
contained in $\pi_1$, and its region type would be
\C{Pair<\ObjZ,\ObjZ><$\pi_1$,$\pi_1$,$\pi_1$>}, which we abbreviate as
\C{Pair<\ObjZ,\ObjZ>@$\pi_1$}. In general, we treat
$B\inang{\tbar}@\pi$ as being equivalent to
$B\inang{\tbar}\inang{\bar{\pi}}$. Region annotation on type $a$,
where $a$ is a type variable, assumes the form $a@\pi$. If $a$ is
instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
\C{Pair} object contained in  $\pi$.

Classes in \FB are independently parameterized over types and regions.
While this design decision has a downside in that it allows type
variables to only denote the types of objects contained in a single
region, it yields benefits that outweigh the costs. In particular, it
lets us support region-polymorphic higher-order functions as class
fields.  This allows, for example, a generic class, whose type
parameters are $a$ and $b$, to contain a region-polymorphic function
of type $\inang{\rho_0,\rho_1,\rho_2 \,|\, \rho_1 \outlives \rho_0
\conj \rho_2 \outlives \rho_0}(a@\rho_1,b@\rho_2) \rightarrow
\C{Pair}\inang{a,b}\inang{\rho_0,\rho_1,\rho_2}$ as its field. Such
region-polymorphic higher-order fields are used frequently by the
dataflow operators, which apply them in the context of various regions
(\emph{e.g.,} see Fig.~\ref{fig:motivating-eg-in-broom}). Keeping type
and region parameterizations separate also simplifies the type system
so that inference becomes practical (Sec.~\ref{sec:type-inference}).
The need for polymorphism at the field level is also why \FB treats
function closures specially, rather than as objects of type \C{Func}
as in C\#.

@@

Note that mutable object fields are conspicuously absent from the \FB
model (and also the FGJ model in general), but this isn't a major
shortcoming considering that constructor application, which happens
during a new object creation, includes assignments to the object
fields (see FGJ~\cite{fgj}). Thus the type system is already obligated
to handle unsafe assignments.

@@

The operational semantics of \FB describes a non-trivial runtime
component that introduces memory regions and performs run-time
verification of their typestate. Fig.~\ref{fig:fb-syntax-extra} shows the
additional language constructs of \FB that manifest only at run-time.
A location ($\loc$) abstracts all the memory locations associated with
a region (i.e., each memory region is associated with a single
location). A transferable region can be in one of three possible
states: closed ($\USED$), open/live ($\LIVE$), and transferred/freed
($\XFERRED$). Constructs \C{letd} and \C{opened} are the run-time
manifestations of \C{letregion} and \C{open}; \C{letregion} reduces to
\C{letd} while allocating a (static) region, and \C{open} reduces to
\C{opened} while opening a (transferable) region. An \C{opened}
expression is tagged with the typestate ($s$) of the transferable
region before it is opened (as Fig.~\ref{fig:region-fsm} shows, a
transferable region can be \C{open}'d from either an open state or a
closed state). Special value $\bot$ denotes an exception.
